The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-01 kalman_walk_paper.tex
general edits to paper

Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
 % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
%
%AND THESE?
  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.

@@

\begin{example}[All Phenotypically Equivalent Oscillators] \label{ex:all_osc}
The oscillator of example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
by an invertible matrix $V$.
If we further require $B$ and $C$ to be invariant then we need $VB=B$ and $CV=C$.
Solving these equations,
we find that a one-parameter family $(A(\tau), B, C)$ describes the set of all two-gene systems
phenotypically equivalent to the oscillator,
where
    \begin{align*}
      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \text{for} \ \tau \neq 1 .
    \end{align*}
The resulting set of systems,
and their dynamics, are depicted in Figure \ref{fig:all_osc}
\end{example}

@@

Is there evidence that this is actually occurring?
System drift and network rewiring has been inferred across the tree of life \citep{dalal2017transcription, johnson2017rewiring},
and there is often significant regulatory variation segregating within populations \citep{XXX}.
\js{cite sergey?}
Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}.
Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),
%strongly reproductively incompatible species
genetically distinct species
may be nearly morphologically indistinguishable.
\plr{check the skinks have RI!!} \jss{the paper just says highly ``genetically distinct''; nothing is stated about RI}
%Of course, selective pressures have likely not been constant over evolutionary time
%in any of these examples,
%but the process we describe here will still occur.
%This model therefore may provide a neutral ``null'' model of gene network evolution, a strategy advocated for by \citet{koonin2016splendor}.
\plr{revisit}
Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for in \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}). If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.

@@

and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch_force} 

>>>>

Although nonidentifiability may frustrate the occasional engineer or scientist, viewed from another angle,
this concept can provide a starting point for thinking about externally equivalent systems
-- systems that evolution can explore, so long as the parameters and structures can be realized biologically.
These functional symmetries manifest in convergent and parallel evolution, 
as well as \emph{developmental system drift}: the observation that
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{kimura1981possibility, true2001developmental, tanay2005conservation, tsong2006evolution, hare2008sepsid, lavoie2010evolutionary, vierstra2014mouse, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription}.
 % \jss{should hare2008sepsid be here? It's sequence but not ``system'' drift}
% are these good here? {taylor2016diverse -- maybe not, matsui2015regulatory -- I think this relevant, stergachis2014conservation}.
%
%AND THESE?
  %% Over the last several years, different computational models have been applied to study reproductive incompatibility and speciation in models of gene regulation \citep{porter2002speciation, tulchinsky, palmer2009dynamics, tulchinsky2014hybrid, khatari2015simple}.
  %\citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybrid incompatibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproductive isolation primarily following directional selection for a change in expression levels in each allopatric population; the evidence for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
%Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overall network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.

@@

\begin{example}[All phenotypically equivalent oscillators] \label{ex:all_osc}
The oscillator of example \ref{ex:oscillator} is minimal, and so any equivalent system is a change of coordinates
by an invertible matrix $V$.
If we further require $B$ and $C$ to be invariant then we need $VB=B$ and $CV=C$.
Solving these equations,
we find that a one-parameter family $(A(\tau), B, C)$ describes the set of all two-gene systems
phenotypically equivalent to the oscillator,
where
    \begin{align*}
      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \text{for} \ \tau \neq 1 .
    \end{align*}
The resulting set of systems,
and their dynamics, are depicted in Figure \ref{fig:all_osc}
\end{example}

@@

Is there evidence that this is actually occurring?
System drift and network rewiring has been inferred across the tree of life \citep{dalal2017transcription, johnson2017rewiring},
and there is often significant regulatory variation segregating within populations \citep{XXX}.
Transcription in hybrids between closely related species with conserved transcriptional patterns can also be divergent 
\citep{haerty2006gene, maheshwari2012cis, coolon2014tempo, michalak2004association}.
Furthermore, in cryptic species complexes (e.g., sun skinks \citep{barley2013challenge}),
%strongly reproductively incompatible species
genetically distinct species
may be nearly morphologically indistinguishable.
\plr{check the skinks have RI!!} \jss{the paper just says highly ``genetically distinct''; nothing is stated about RI}
%Of course, selective pressures have likely not been constant over evolutionary time
%in any of these examples,
%but the process we describe here will still occur.
%This model therefore may provide a neutral ``null'' model of gene network evolution, a strategy advocated for by \citet{koonin2016splendor}.
\plr{revisit}
Of course the basic assumptions of this model will be violated in practice (constant selective pressures, etc.), however, this model can function as a ``neutral null'' description of gene network evolution (a strategy advocated for by \citet{lynch2007frailty, fay2008evaluating, koonin2016splendor}). If this model succeeds in describing actual system evolution, it can be inferred that the mechanism underlying species formation does not require the inclusion of adaptation or changes in selection to explain the emergence of hybrid incompatibility.

@@

and plausible mechanisms for the origin of Dobzhansky--Muller incompatibilities by neutral genetic drift have been proposed \citep{lynch2000origin} 
